\Large Last Writer Slices and Communication Traps

### 2013-11-06 paper.tex
committing more changes in eval.  got rid of redundancy stuff

By comparing \ctraps and last writer slicing, we see four applications ({\tt
dedup}, {\tt streamcluster}, {\tt fluidanimate}, {\tt ferret}) that have
\ctraps overhead that is probably too high for always-on use (averaging 153\%)
drop to a last writer slicing overhead that is probably low enough for always-on use
(averaging 38\%).  The large decrease indicates these programs perform
relatively more reads than other applications, and thus benefit more from
eliminating read instrumentation.
%\paragraph{\ctrapsmm}
%\ctrapsmm has very similar overheads to \ctrapsfull in most cases.   The only case
%with a major difference was {\tt vips}, indicating it executes more redundant
%reads than the other applications.  The similarity in the two designs'
%overheads is expected because our default ``no-op'' read instrumentation does
%not contribute much to overheads in most cases.  Hence, unless \ctrapsmm culls
%calls with high dynamic frequency, it is unlikely to reduce overheads much.  We
%show in Section~\ref{sec:appperf} that \ctrapsmm provides more performance
%benefit when used to implement applications that make read instrumentation sites
%more costly.
\\
\\
These high-level results support our main claim that the runtime overhead of
\ctraps is low enough for deployed systems for many 
applications, especially for servers.  We now further analyze our observed overheads, illuminating
differences between the lower-overhead server programs and the higher-overhead
PARSEC programs.

@@

\subsubsection{Overheads Due to Conservative Escape Analysis}
\label{sec:eval:conservative}
{\tt swaptions} had the highest overheads in our tests.  We investigated their
cause and determined they are mostly artificial and due to shortcomings of our
escape analysis.  We located the inner-loop function {\tt
HJM\_SimPath\_Forward\_Blocking}.  The function uses two local matrices that
are allocated through an external call.  Our escape analysis cannot determine
that the call only does allocation and that the matrices are local, so it
conservatively assumes the matrices escape.  We manually inlined the allocation
(and a matching deallocation), and escape analysis eliminated their
instrumentation.  The program's overhead dropped from about 420\% overhead for
\ctraps to about 300\% overhead; the change for last writer slicing was from 300\% to about
130\% overhead.  We also looked at an output matrix passed into the inner-loop
function.  The matrix is allocated and deallocated in the inner-loop's caller,
but is never shared.  Eliminating instrumentation on accesses to this matrix
yielded an overhead of about 46\% for \ctraps, and to just 30\% for
last writer slicing.

Note that both these changes amount to inlining and preserve the semantics and
sharing pattern of the program.  We expect that more sophisticated escape
analysis would find these local accesses and achieve these performance gains
automatically.

\subsubsection{PARSEC vs. Servers}
\label{sec:eval:parsecserver}
In our tests, server programs were better able to tolerate the overhead of
last writer slicing and \ctraps.   We discern three reasons for this difference.

\paragraph{Independent Parallelism}
In most cases, servers have more independent parallel work.  Server programs
set up a pool of threads to handle requests.  Each thread incurs some
instrumentation latency, but that latency is hidden by other threads making
progress on independent requests.  This analysis also applies to the PARSEC
applications that had performance similar to the servers.  For instance, {\tt
blackscholes} uses largely independent threads to compute on different regions
of a matrix.  In addition, threads performing independent computation ({\em
i.e.}, non-communicating computation) are likely to operate mostly on
non-escaping, local data.  Local data are not instrumented by our compiler,
sparing the overhead.  We characterize the amount of local data in each
application in detail in Section~\ref{sec:char}.  We show that the server
programs have a larger proportion of provably local accesses than PARSEC.  The
difference supports the fact that the servers have more independent work ({\em
i.e.}, more local accesses) than PARSEC.

\paragraph{Latency Hiding I/O}
The servers perform more I/O.  Similarly to the effect of abundant parallelism,
I/O helps hide instrumentation latency.  While I/O is naturally part of many
applications, we aimed to minimize the effect of I/O in our experiments.  For
MySQL and Apache, we used local disks for logs and resources, and connected via
a local socket, rather than using the network.  In spite of these controls, I/O
remains, and hides latency in our server applications.  LevelDB's benchmark is
designed to avoid all I/O, creating and manipulating a database entirely in a
single process.  The absence of I/O may account for some of the difference in
overhead between LevelDB and the other servers.  Note that the effects of I/O
are not merely experimental noise -- in real deployed software the latency
hiding effects of I/O are likely to be even {\em more} pronounced.

\paragraph{Interference with Optimization}
The PARSEC programs are more heavily hand-optimized. In several cases
resorting to hand-coded assembly and cache-aware memory access patterns.
Running instrumentation that accesses the \lwt alongside these computations may
disrupt their cache performance, or other delicately optimized behavior,
leading to higher overheads.

Such carefully optimized code is likely to have been written by expert
programmers.  As with other code written by experts ({\em e.g.}, lock-free
data-structures), this code may not require as much {\em in situ} analysis
support as code written by distributed teams of open source developers or
novice programmers.  Depending on the maturity and level of verification of a
piece of carefully hand-optimized code, it might make sense to elide all
communication tracking instrumentation to preserve performance.

%\subsubsection{Breaking Down the Design Space}
%Figure~\ref{fig:perfall} also shows the impact of context-sensitivity, lossy
%optimization, and sampling on our overheads.  
%
%\paragraph{Context Sensitivity}
%Context-sensitivity imposes a substantial overhead in some cases.  The
%additional cost of context-sensitivity is about 10\% of the total overhead for
%PARSEC.  For server applications, the cost is less substantial, but still a
%non-negligible 4\% over the baseline.  The cost of context-sensitivity comes
%from the need to compute and store multiple addresses, rather than a single
%program counter.  
%
%In the context insensitive case, updating the \lwt requires a single register
%read ({\em rip}) a bitwise {\tt or} to combine the thread identifier and the
%instruction pointer, and the actual write to update the \lwt entry.  In the
%context sensitive case, updating the \lwt requires looking back over the call
%stack, and a bitwise {\tt or} operation at each return address.  The cost is at
%least a multiple of the cost of the constext insensitive case.  However, to
%restrict overhead to write instrumentation calls, we walk stack frames, rather
%than iteratively computing call stacks at each call and return.  As a result,
%the cost is amplified by the need to walk the stack to find return addresses.
%Future work could reduce this cost using Breadcrumbs~\cite{breadcrumbs} or
%Probabilistic Calling Contexts~\cite{probcallcont}. \addtodo{Use the Aviso
%iterative context tracker -- maybe it's actually faster?}

%\paragraph{Redundant Read Optimization}
%
%Our static analysis to remove read instrumentation reduces performance
%overheads for PARSEC programs, but does not provide much benefit in server
%programs.  In particular, removing redundant read instrumentation benefits {\tt
%x264} and {\tt swaptions}.  In the case of {\tt swaptions}, the gains are due
%to a \addtodo{Read Swaptions Code -- where's the beef?}.  In the case of {\tt
%x264}, the improvement is due to \addtodo{What is it?}.
%
%Note that the experiments depticted in Figure~\ref{fig:perfall} reflect our
%runtime in its base configuration -- writes update the \lwt, and reads only
%call a {\em no-op} hook.  Our base configuration targets manual debugging,
%which does not require actions on reads.  In more sophisticated applications of
%\ctraps, such as communication graph collection, action is required on reads.
%In Section~\ref{sec:graphperf}, we show that this optimization more
%substantially improves performance when read instrumentation sites do more
%work.
%
%\paragraph{Sampling}
%Sampling reduces overheads for both server and PARSEC applications.
%\addtodo{Explain why sampling is awesome sometimes and less awesome other
%times}

%\paragraph{Annotating Away Instrumentation}
%\addtodo{Discuss the /*Sequential*/ region in streamcluster and how not instrumenting those loads and stores might increase perf a lot}

%blackscholes 2 4
%canneal 124 226
%dedup 1 171
%ferret 40 792
%fluidanimate 3 149
%streamcluster 22 47
%swaptions 2 19
%vips 1919 9116
%x264 2987 5960

%blackscholes 3.77358490566038 7.54716981132075
%canneal 15.7360406091371 28.6802030456853
%dedup 0.128865979381443 22.0360824742268
%ferret 1.0584810796507 20.9314633500926
%fluidanimate 0.398406374501992 19.6547144754316
%streamcluster 5.44554455445545 11.6336633663366
%swaptions 0.87719298245614 8.33333333333333

\subsection{Context-Sensitivity}
We briefly evaluated the cost of context-sensitivity for \ctraps.  With a
bounded 2-address context, the average overhead for server programs was 18\%
and the average overhead for PARSEC was 145\%.  These data show that a
restricted form of context-sensitive analysis is possible in \ctraps with
overheads low enough for production.  This possibility broadens the scope of
applicability for \ctraps.  We omit a full analysis of these results here due
to space constraints.

\subsection{Evaluating CTraps Applications}
\label{sec:appperf}
We evaluated our CCI-Prev and communication graph collection CTraps
applications.  Our results show that these CTraps applications have modest
overheads that scale with their analysis complexity.  In several cases -- most
notably, in tests with our server benchmarks -- their overhead is low enough
for production use.  We also show that our redundancy analysis saves some
performance overhead in some cases.

@@

\subsection{Characterizing \ctraps}
\label{sec:char}
We characterized the instrumentation pruned by escape and redundancy analysis.
We report our results in Table~\ref{tab:char}.  Each column shows the percent
of instrumentation sites pruned and the absolute number of sites in
parenthesis.  The first column shows the amount of static instrumentation
pruned by escape analysis.  The second column shows the amount of static
instrumentation pruned by our redundancy analysis.  The third column shows how
many dynamic instances of instrumentation were eliminated by redundancy analysis.  To
collect the data in the third column, we implemented a special instrumentation
pass that marked pruned functions statically so they could be counted
separately from non-pruned functions at runtime.

@@

The data also show that fewer accesses were eliminated by redundancy analysis
than by escape analysis.  In some cases ({\em e.g.}, {\tt swaptions}), the
fraction of static and dynamic instrumentation sites eliminated is small.  In
these cases, the few pruned calls were infrequently executed and as
Figure~\ref{fig:perfapps} shows, there was little reduction in overhead.  In
other cases, like {\tt x264}, the fraction of calls pruned was substantial 
and yielded the largest performance boost.

>>>>

By comparing \ctraps and last writer slicing, we see four applications ({\tt
dedup}, {\tt streamcluster}, {\tt fluidanimate}, {\tt ferret}) that have
\ctraps overhead that is probably too high for always-on use (averaging 153\%)
drop to a last writer slicing overhead that is probably low enough for
always-on use (averaging 38\%).  The large decrease indicates these programs
perform relatively more reads than other applications, and thus benefit more
from eliminating read instrumentation.

These high-level results support our main claim that the runtime overhead of
\ctraps is low enough for deployed systems for many applications, especially
for servers.  We now further analyze our observed overheads, illuminating
differences between the lower-overhead server programs and the higher-overhead
PARSEC programs.

@@

\subsubsection{Overheads Due to Conservative Escape Analysis}
\label{sec:eval:conservative}
{\tt swaptions} had the highest run time overhead in our tests.  We
investigated the cause and determined it is largely due to shortcomings in our
escape analysis.  We located the inner-loop function {\tt
HJM\_SimPath\_Forward\_Blocking}.  The function uses two local matrices that
are allocated through an external call.  Our escape analysis cannot determine
that the call only does allocation and that the matrices are local, so it
conservatively assumes the matrices escape.  We manually inlined the allocation
(and a matching deallocation), and escape analysis eliminated their
instrumentation.  The program's overhead dropped from about 420\% overhead for
\ctraps to about 300\% overhead; the change for last writer slicing was from
300\% to about 130\% overhead.  We also looked at an output matrix passed into
the inner-loop function.  The matrix is allocated and deallocated in the
inner-loop's caller, but is never shared.  Eliminating instrumentation on
accesses to this matrix yielded an overhead of about 46\% for \ctraps, and to
just 30\% for last writer slicing.

Note that both these changes amount to inlining and preserve the semantics and
sharing pattern of the program.  We expect a more sophisticated escape
analysis would find these local accesses and achieve these performance gains
automatically.

\subsubsection{PARSEC vs. Servers}
\label{sec:eval:parsecserver}
In our tests, server programs were better able to tolerate the overhead of
last writer slices and \ctraps.   We discern three reasons for this difference.

\paragraph{Independent Parallelism}
In most cases, the servers we evaluated have abundant, completely independent
parallel work.  Server programs use a pool of threads to handle requests.
Each thread incurs some instrumentation latency, but that latency is hidden by
other threads making progress on independent requests.  This analysis also
applies to the PARSEC applications that had performance similar to the servers.
For instance, {\tt blackscholes} uses largely independent threads to compute on
different regions of a matrix.  In addition, threads performing independent
computation ({\em i.e.}, non-communicating computation) are likely to operate
mostly on non-escaping, local data.  Local data are not instrumented by our
compiler, sparing the overhead.  We characterize the amount of local data in
each application in detail in Section~\ref{sec:char}.  We show that the server
programs have a larger proportion of provably local accesses than PARSEC.  The
difference supports the fact that the servers have more independent work ({\em
i.e.}, more local accesses) than PARSEC.

\paragraph{Latency Hiding I/O}
The servers perform more I/O.  Similarly to the effect of abundant parallelism,
I/O helps hide instrumentation latency.  While I/O is naturally part of many
applications, we aimed to minimize the effect of I/O in our experiments.  For
MySQL and Apache, we used local disks for logs and resources, and connected via
a local socket, rather than using the network.  In spite of these controls, I/O
remains, and hides added instrumentation latency in our server applications.
LevelDB's benchmark is designed to avoid all I/O, creating and manipulating an
in-memory database in a single process.  The absence of I/O may account for some
of the difference in overhead between LevelDB and the other servers.  Note that
the effects of I/O are not merely experimental noise -- in real deployed
software the latency hiding effects of I/O are likely to be even {\em more}
pronounced, favoring the use of our system.

\paragraph{Interference with Optimization}
The PARSEC programs are heavily hand-optimized. In several cases
resorting to hand-coded assembly and cache-aware memory access patterns.
Running instrumentation that accesses the \lwt alongside these computations may
disrupt their cache performance, or other delicately optimized behavior,
leading to higher overheads.

Such carefully optimized code is likely to have been written by expert
programmers.  As with other code written by experts ({\em e.g.}, lock-free
data-structures), this code may not require as much {\em in situ} analysis
support as code written by distributed teams of open source developers or
novice programmers.  Depending on the maturity and level of verification of a
piece of carefully hand-optimized code, it might make sense to elide 
last writer slicing and \ctraps code to preserve performance.

\subsection{Context-Sensitivity}
{\bf TODO: Was this LWS or CTRAPS-NOOP or what?}
We briefly evaluated the cost of context-sensitivity for \ctraps.  With a
bounded 2-address context, the average overhead for server programs was 18\%
and the average overhead for PARSEC was 145\%.  These data show that a
restricted form of context-sensitive analysis is possible in \ctraps with
overheads low enough for production.

\subsection{Evaluating CTraps Applications}
\label{sec:appperf}
We evaluated our CCI-Prev and communication graph collection CTraps
applications.  Our results show that these CTraps applications have modest
overheads that scale with their analysis complexity.  In several cases -- most
notably, in tests with our server benchmarks -- their overhead is low enough
for production use.

@@

\subsection{Characterizing \ctraps}
\label{sec:char}
We characterized the instrumentation pruned by escape analysis.  We report our
results in Table~\ref{tab:char}.  The table shows the percent of
instrumentation sites pruned and the absolute number of sites in parenthesis.

@@
